---
title: "Lab_1"
author: "Vaibhav Rastogi"
date: "April 5, 2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) 
library(lubridate)
```

# Basic Descriptive Analytics with R

### Getting Data
* Open Lab1 and unzip. You should have the following files:
+ IBMStock.csv, GEStock.csv, ProcterGambleStock.csv, CocaColaStock.csv, BoeingStock.csv.
+ Note: don't open the CSVs in Excel before importing to R as it can mess up the Date field

### Importing Data into R
* Import each CSV into R as a dataframe using the read.csv() function
* When you import the data, give each dataframe the following names:
+ "IBM", "GE", "ProcterGamble", "CocaCola", and "Boeing"
```{r read, prompt=TRUE}
IBM = read.csv('IBMStock.csv')
GE = read.csv('GEStock.csv')
ProcterGamble = read.csv('ProcterGambleStock.csv')
CocaCola = read.csv('CocaColaStock.csv')
Boeing = read.csv('BoeingStock.csv')
```

* Each dataframe has two columns:
+ Date: the date of the stock price, always given as the first of the month.
+ StockPrice: the average stock price of the company in the given month.

### Clean Date Field
* You'll notice the dates will not be in a format R can understand right after you import. Check this out using the str(). What is the data type of the Data variable? ANSWER:
```{r structure, prompt=TRUE}
str(IBM$Date)
str(GE$Date)
str(ProcterGamble$Date)
str(CocaCola$Date)
str(Boeing$Date)
```
**the Date field is a Factor data type**   

* We can use the as.Date() to convert the dates. Here's an example of applying for the GE dataframe:
$ GE$Date = as.Date(GE$Date, "%m/%d/%y") 
+ Using the sample code above, convert the datatype of the date field for all 4 files.
```{r convert to date, prompt=TRUE}
IBM$Date = as.Date(IBM$Date, "%m/%d/%y")
GE$Date = as.Date(GE$Date, "%m/%d/%y")
ProcterGamble$Date = as.Date(ProcterGamble$Date, "%m/%d/%y")
CocaCola$Date = as.Date(CocaCola$Date, "%m/%d/%y")
Boeing$Date = as.Date(Boeing$Date, "%m/%d/%y")
```
\pagebreak  

### Warm-up/Basic statistics Questions:
1. How many rows of data are in each dataset?
```{r rows, prompt=TRUE}
nrow(IBM)
nrow(GE)
nrow(ProcterGamble)
nrow(CocaCola)
nrow(Boeing)
```
**each data set has 480 rows**   

2. What is the earliest/latest year in our datasets?  

Stock | Earliest Year | Latest Year
----- | ------------- | ------------
IBM | `r min(year(IBM$Date))` | `r max(year(IBM$Date))`
GE | `r min(year(GE$Date))` | `r max(year(GE$Date))`
ProcterGamble | `r min(year(ProcterGamble$Date))` | `r max(year(ProcterGamble$Date))`
CocaCola | `r min(year(CocaCola$Date))` | `r max(year(CocaCola$Date))`
Boeing | `r min(year(Boeing$Date))` | `r max(year(Boeing$Date))`

3. For the period above what is the average stock price of Coca Cola?

**Average stock price of Coca Cola: `r mean(CocaCola$StockPrice)`**   

4. What is the maximum price of IBM during this period?

**Maximum stock price of IBM: `r max(IBM$StockPrice)`**   

5. What is the standard deviation of P&G stock price over this period?

**Standard deviation of P&G stock price over the period: `r sd(ProcterGamble$StockPrice)`**   

6. What is the median price of Boeing in the last 5 years for which we have data?
```{r median price for last 5 years, prompt=TRUE}
(Boeing5YearMedianStockPrice <- Boeing %>%
  mutate(Year = year(Date)) %>%
  filter(Year >= (max(year(Date)) - 4)) %>%
  summarise(MedianStockPrice = median(StockPrice)))
```
**Median stock price of Boeing in the last 5 years for which we have data: `r Boeing5YearMedianStockPrice`**  
\pagebreak  

### Basic Plotting Questions
**Part 1:** Plot the StockPrice of Coca-Cola on the Y-axis across Date on the x-axis using the basic plot() function.
```{r plot StockPrice vs Date, prompt=TRUE}
plot(CocaCola)
```

What do you see when you use the default plot function what do you see? Scatter-plot, eh? If you want a line graph instead, add the argument type="l" to your plot function.
```{r plot StockPrice vs Date line graph, prompt=TRUE}
plot(CocaCola, type= 'l')
```

**Answer the following questions:**   
1. Identify the year during which Coca-Cola had the highest/lowest stock price?
```{r date of highest stock price, prompt=TRUE}
(CocaColaHigh <- filter(CocaCola, CocaCola$StockPrice == max(CocaCola$StockPrice)))
```
**The highest stock price was on `r CocaColaHigh$Date` and the price was `r CocaColaHigh$StockPrice`**   

```{r date of lowest stock price, prompt=TRUE}
(CocaColaLow <- filter(CocaCola, CocaCola$StockPrice == min(CocaCola$StockPrice)))
```
**The lowest stock price was on `r CocaColaLow$Date` and the price was `r CocaColaLow$StockPrice`**   

2. What calendar year did it look to have the biggest (Year-over-Year) percentage increase?
```{r biggest year-over-year percentage increase, prompt=TRUE}
CocaColaYearly <- CocaCola %>%
  mutate(Year = year(Date)) %>%
  group_by (Year) %>%
  summarize (LastStockPrice = last(StockPrice)) %>%
  mutate(StockPriceDiff = c(NA,diff(LastStockPrice)), 
         PrevLastStockPrice = LastStockPrice - c(NA,diff(LastStockPrice)),
         PercentDiff = ((c(NA,diff(LastStockPrice)))/(LastStockPrice - c(NA,diff(LastStockPrice)))))
(CocaColaYearlyMax <- CocaColaYearly %>%
    filter(near(CocaColaYearly$PercentDiff,max(CocaColaYearly$PercentDiff,na.rm = TRUE))))

```
**The calendar year with the biggest percent increase (Year-to-Year) is `r CocaColaYearlyMax$Year` with an increase of `r 100 * CocaColaYearlyMax$PercentDiff`%**   

**Part 2:** Next we want to add P&G stock price onto the same graph. If you keep your plot window open
and type in: lines(ProcterGamble$Date, ProcterGamble$StockPrice) you should see your plot update.
Go back to the plot function and add argument col="red" and col="blue" argument for CocaCola and
P&G respectively.
```{r plot CocaCola and ProcterGamble, prompt=TRUE}
plot(CocaCola,col='red',type = 'l') 
lines(ProcterGamble,col = 'blue')
```

**Answer the following questions:**  
1. In March of 2000 the stock market plummeted as the tech bubble burst. 
Using the plot above, which company's stock dropped more (relatively - i.e. percentage-wise)?
```{r March 2000, prompt=TRUE}
CocaColaMar <- filter(CocaCola, between(Date, as.Date("2000-02-01"), as.Date("2000-04-01")))
ProcterGambleMar <- filter(ProcterGamble, between(Date, as.Date("2000-02-01"), as.Date("2000-04-01")))
plot(CocaColaMar,col='red',type = 'l', ylim=c(0, 100))
lines(ProcterGambleMar, col = 'blue')
CocaColaDecr <- last(CocaColaMar$StockPrice) - first(CocaColaMar$StockPrice)
CocaColaPercDecr <- CocaColaDecr / first(CocaColaMar$StockPrice)
ProcterGambleDecr <- last(ProcterGambleMar$StockPrice) - first(ProcterGambleMar$StockPrice)
ProcterGamblePercDecr <- ProcterGambleDecr / first(ProcterGambleMar$StockPrice)
if (CocaColaDecr < ProcterGambleDecr) { 
  RelativeDecr = 'CocaCola' } else { 
  RelativeDecr = 'ProcterGamble' }
if (CocaColaPercDecr < ProcterGamblePercDecr) { 
  PercentDecr = 'CocaCola' } else { 
  PercentDecr = 'ProcterGamble' } 
```
Stock | Relative Decrease | Percent-wise Decrease
----- | ----------------- | ---------------------
CocaCola | `r CocaColaDecr` | `r CocaColaPercDecr`
ProcterGamble | `r ProcterGambleDecr` | `r ProcterGamblePercDecr`

**`r RelativeDecr` had a greater relative decrease in March 2000. **
**`r PercentDecr` had a greater percent-wise decrease in March 2000. **

2. In the year 1983 which company stock was going up? Which was going down?
```{r 1983, prompt=TRUE}
CocaCola1983 <- 
  filter(CocaCola, CocaCola$Date == '1983-12-01')[1,'StockPrice'] - 
  filter(CocaCola, CocaCola$Date == '1983-01-01')[1,'StockPrice']
ProcterGamble1983 <- 
  filter(ProcterGamble, ProcterGamble$Date == '1983-12-01')[1,'StockPrice'] - 
  filter(ProcterGamble, ProcterGamble$Date == '1983-01-01')[1,'StockPrice']
```
**In 1983, CocaCola stock price was `r if (CocaCola1983 > 0) {'going up'} else {'going down'}` **
**In 1983, ProcterGamble stock price was `r if (ProcterGamble1983 > 0) {'going up'} else {'going down'}` **

3. Across the entire time period shown in your plot which stock had a generally lower price?
```{r average, prompt=TRUE}
CocaColaMean <- mean(CocaCola$StockPrice)
ProcterGambleMean <- mean(ProcterGamble$StockPrice)
```
**Over the entire time period `r if (CocaColaMean < ProcterGambleMean) {'CocaCola'} else {'ProcterGamble'}` generally had a lower stock price**   
\pagebreak  

### Data Visualization from 1995-2005:
Instead of looking at the plot across the entire date range, we want to see what's happening between 1995-2005. Remember, you can use the matrix notation [rows, columns] to subset data.
Keeping that in mind what row numbers represent the Dates from 1995-2005?
```{r 1995-2005 rows, prompt=TRUE}
(StartRow <- which(CocaCola$Date == as.Date('1995-01-01')))
(EndRow <- which(CocaCola$Date == as.Date('2005-12-01')))
```
**First stock price of the year 1995 sits in row position: `r StartRow`**   
**Last stock price of the year 2005 sits in row position: `r EndRow`**   

Next, run the command below using the date range you found above. 
Note: change the values <Your Answer Start> and <Your Answer End> to your answers above.
```{r 1995-2005 plot, prompt=TRUE}
plot(CocaCola$Date[StartRow : EndRow], CocaCola$StockPrice[StartRow : EndRow], type="l", col="red", ylim=c(0,210))
```

* ylim=c(0,210), makes the y-axis range from 0 to 210.
* type="l" specifies to make it a line graph
* col="red" tell is to make the line colour red
**Answer the following questions:**
1. Which stock price fell the most right after the tech bubble of March 2000?
```{r March 2000 all stocks, prompt=TRUE}
StockPrices <- mutate(IBM,Stock = 'IBM') %>%
  rbind(mutate(GE, Stock = 'GE')) %>%
  rbind(mutate(ProcterGamble, Stock = 'ProcterGamble')) %>%
  rbind(mutate(CocaCola, Stock = 'CocaCola')) %>%
  rbind(mutate(Boeing, Stock = 'Boeing'))
StockPricesMar <- filter(StockPrices, between(Date, as.Date("2000-02-01"), as.Date("2001-04-01")))
ggplot(data = StockPricesMar) + 
  geom_line(mapping = aes(x = Date, y = StockPrice, color = Stock)) +  
  scale_x_date(date_breaks = "1 month", date_labels = "%b-%Y") + 
  theme(axis.text.x = element_text(angle = 90))
```
**In March 2000 ProctorGamble stock prices fell but then a couple months later GE stock prices fell and much more significantly**   

2. What stock had the highest maximum price between 1995-2005?
```{R 1995-2000 max price, prompt=TRUE}
StockPrices95_05 <- filter(StockPrices, between(Date, as.Date("1995-01-01"), as.Date("2005-12-01"))) 
ggplot(data = StockPrices95_05) + 
  geom_line(mapping = aes(x = Date, y = StockPrice, color = Stock)) +  
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") + 
  theme(axis.text.x = element_text(angle = 90))

```
**between 1995 and 2000 IBM had the maximum price of `r max(StockPrices95_05$StockPrice)`**   

3. A few years before the tech bubble of 1997, there was another stock market crash trigged by
economic crisis in Asia in October of 1997. If you compare stock prices from September 1997 to
November 1997, which companies saw a decrease in price? Which company experienced the
biggest decrease?
```{r October 1997, prompt=TRUE}
StockPricesOct <- filter(StockPrices, between(Date, as.Date("1997-09-01"), as.Date("1997-11-01")))
ggplot(data = StockPricesOct) + 
  geom_line(mapping = aes(x = Date, y = StockPrice, color = Stock)) +  
  scale_x_date(date_breaks = "1 month", date_labels = "%b-%Y") + 
  theme(axis.text.x = element_text(angle = 90))
StockPricesOct %>%
  group_by (Stock) %>%
  summarise (Difference = last(StockPrice) - first(StockPrice),
             StartingStockPrice = first(StockPrice),
             EndingStockPrice = last(StockPrice)) %>%
  arrange (Difference)
```
**ProcterGamble saw the biggest drop in stock prices in October 1997 followed by Boeing having a small drop in stock prices.  CocaCola, IBM & GE stock prices increased slightly during that period**   

4. Which stock seemed to provide the best return (i.e. increase in price) between 2004-2005?
```{r 2004-2005, prompt=TRUE}
StockPrices04_05 <- filter(StockPrices, between(Date, as.Date("2004-01-01"), as.Date("2005-12-01")))
ggplot(data = StockPrices04_05) + 
  geom_line(mapping = aes(x = Date, y = StockPrice, color = Stock)) +  
  scale_x_date(date_breaks = "1 month", date_labels = "%b-%Y") + 
  theme(axis.text.x = element_text(angle = 90))
StockPrices04_05 %>%
  group_by (Stock) %>%
  summarise (Difference = last(StockPrice) - first(StockPrice),
             StartingStockPrice = first(StockPrice),
             EndingStockPrice = last(StockPrice),
             PercentDifference = (last(StockPrice) - first(StockPrice))/first(StockPrice)) %>%
  arrange (PercentDifference)
```
**Boeing saw the best return between 2004-2005, both in increase in stock price and percent increase**   

5. Between 1995-2005, which company had the biggest delta between the maximum and
minimum stock price?
```{R 1995-2000 biggest delta, prompt=TRUE}
StockPrices95_05 %>%
  group_by (Stock) %>%
  summarise (
    MinStockPrice = min(StockPrice), 
    MaxStockPrice = max(StockPrice), 
    Difference = max(StockPrice) - min(StockPrice)) %>%
  arrange(Difference)
```
**IBM had the biggest delta between their stock price minimum and stock price maximum**   

6. Which two companies' stock price seem to be the most correlated (i.e. move up/down
together)?  
```{r}
StockPricesFlat <- select(Boeing, BoeingStockPrice = StockPrice) %>%
  cbind(CocaColaStockPrice = CocaCola$StockPrice) %>%
  cbind(GEStockPrice = GE$StockPrice) %>%
  cbind(IBMStockPrice = IBM$StockPrice) %>%
  cbind(ProcterGambleStockPrice = ProcterGamble$StockPrice)
round(cor(StockPricesFlat), 2)
```
**IBM and CocaCola are correlated**
```{r}
ggplot(data = StockPricesFlat) + geom_point(mapping = aes(x=CocaColaStockPrice, y=IBMStockPrice))
StockPrices95_05 %>% 
  filter(Stock == 'CocaCola'| Stock == 'IBM') %>%
  ggplot + geom_line(mapping = aes(x = Date, y = StockPrice, color = Stock)) +  
    scale_x_date(date_breaks = "1 year", date_labels = "%Y") + 
    theme(axis.text.x = element_text(angle = 90))
```
\pagebreak  

### Monthly Trend Analysis
We want to see if there are any monthly patterns (i.e. consistently higher/lower prices at various
months of the year). To do for each company, we essentially want to compare "mean" by month vs the
overall mean across the entire date range. Hint: to sort a dataset by months, you can try use
months(IBM$Date) as the second argument in the tapply() function.
To learn how to use tapply(), read this: https://www.r-bloggers.com/r-function-of-the-day-tapply-2/
**Answer the following questions:**
1. For IBM, compare the average stock price for each month to the its overall average stock price
and identify all the months for which IBM historically had a higher stock price (we call this overindexing)? Which month over-indexed the most?
```{r IBM mean by month, prompt=TRUE}
tapply(IBM$StockPrice, month(IBM$Date), FUN=mean)
mean(IBM$StockPrice)
IBM %>% 
  group_by (Month = month(Date)) %>% 
  summarize(MonthlyAverage = mean(StockPrice)) %>%
  mutate(Difference = MonthlyAverage - mean(IBM$StockPrice)) %>%
  ggplot + geom_line(mapping = aes(x = Month, y = Difference))
```
**Jan, Feb, Mar, Apr, May have average (mean) stock prices greater than the average stock price**   
**Feb is over-indexed the most**   

2. Repeat the tapply() function you used to solve the last question for each of the 4 remaining
companies. Do any of two or more companies have their highest stock price in the same months
as each other? Which companies and months does this happen for?
```{r mean by month, prompt=TRUE}
tapply(GE$StockPrice, month(GE$Date), FUN=mean)
mean(GE$StockPrice)
tapply(ProcterGamble$StockPrice, month(ProcterGamble$Date), FUN=mean)
mean(ProcterGamble$StockPrice)
tapply(CocaCola$StockPrice, month(CocaCola$Date), FUN=mean)
mean(CocaCola$StockPrice)
tapply(Boeing$StockPrice, month(Boeing$Date), FUN=mean)
mean(Boeing$StockPrice)
IBMSummary <- IBM %>% 
  group_by (Month = month(Date)) %>% 
  summarize(MonthlyAverage = mean(StockPrice)) %>%
  mutate(Difference = MonthlyAverage - mean(IBM$StockPrice), Stock = 'IBM')
GESummary <- GE %>% 
  group_by (Month = month(Date)) %>% 
  summarize(MonthlyAverage = mean(StockPrice)) %>%
  mutate(Difference = MonthlyAverage - mean(GE$StockPrice), Stock = 'GE')
ProcterGambleSummary <- ProcterGamble %>% 
  group_by (Month = month(Date)) %>% 
  summarize(MonthlyAverage = mean(StockPrice)) %>%
  mutate(Difference = MonthlyAverage - mean(ProcterGamble$StockPrice), Stock = 'ProcterGamble')
CocaColaSummary <- CocaCola %>% 
  group_by (Month = month(Date)) %>% 
  summarize(MonthlyAverage = mean(StockPrice)) %>%
  mutate(Difference = MonthlyAverage - mean(CocaCola$StockPrice), Stock = 'CocaCola')
BoeingSummary <- Boeing %>% 
  group_by (Month = month(Date)) %>% 
  summarize(MonthlyAverage = mean(StockPrice)) %>%
  mutate(Difference = MonthlyAverage - mean(Boeing$StockPrice), Stock = 'Boeing')
StockSummary <- IBMSummary %>%
  rbind(GESummary) %>%
  rbind(ProcterGambleSummary) %>%
  rbind(CocaColaSummary) %>%
  rbind(BoeingSummary)
filter(StockSummary,Difference > 0)
ggplot(data = StockSummary) + 
  geom_line(mapping = aes(x = Month, y = Difference, color = Stock))
```
**IBM - Jan, Feb, Mar, Apr, May**   
**GE - Jan, Feb, Mar, Apr, May**   
**ProctorGamble - Jan, Feb, May, Nov, Dec**   
**CocaCola - Jan, Feb, Mar, Apr, May, Jun**   
**Boeing - Feb, Mar, Apr, May, Jun, Aug**   
**IBM & GE are higher than average on the same months - Jan, Feb, Mar, Apr, May**   
**all Stocks are higher than average in Feb & May**   
**all Stocks are lower than average in Sep & Oct**   

3. What trend do you see for the months of December vs January for each company? Is there an
over-arching trend that applies to all companies when comparing all historical December vs
January stock prices?  

**January has a higher average Stock Price than December for each Stock**   
**according to this website https://www.investopedia.com/terms/j/januaryeffect.asp there is something called the December Selloff and the January Effect that causes prices to go up in January over December**   
**but according to this website http://www.moneychimp.com/features/monthly_returns.htm that has the S&P 500 since 1950, the December Selloff and the January Effect is a myth although September is a particularly bad month**   
